Open-vocabulary semantic exploration requires mobile robots to search for open-language targets in previously unseen environments, while coping with partial observability, perceptual ambiguity, and limited onboard compute. Existing approaches either rely on reinforcement learning with extensive training in simulation, or use zero-shot vision-language models without persistent memory, which can lead to inefficient exploration and brittle decisions under noisy observations. This thesis presents \textbf{SAGE} (Semantic-Aware Guided Exploration), a hybrid semantic exploration framework that combines (i) frontier-based semantic exploration driven by vision-language similarity, (ii) persistent three-dimensional semantic memory based on OpenFusion for long-horizon exploitation, and (iii) multi-source semantic fusion for robust target confirmation. The method integrates geometric feasibility and navigational constraints into frontier scoring and fuses detector confidence, vision-language similarity, and memory evidence using a Noisy-Or formulation to reduce missed detections under conservative decision thresholds.

\textbf{SAGE} is evaluated in a photorealistic IsaacSim setup on Habitat-Matterport 3D Version 2 scenes using a multi-object search protocol. Experiments quantify (1) end-to-end navigation performance against representative baselines, (2) sensitivity to the exploration-memory weighting, (3) robustness to semantic map granularity via the OpenFusion top-$k$ retrieval depth, and (4) the effect of multi-source fusion on detection precision-recall characteristics. Results show that, within this custom evaluation setting, \textbf{SAGE} achieves the highest Success weighted by Path Length among the compared methods while maintaining a competitive Success Rate. Further analyses show that intermediate exploration-memory weightings yield the most stable trade-off between reactivity and exploitation, and that balanced exploration mitigates the impact of noisy or overly dense semantic maps. Multi-source fusion substantially reduces false negatives at conservative operating points while maintaining precision, supporting a two-stage verification strategy in the behavior tree. Finally, a deployment-oriented analysis reports GPU memory consumption and multi-rate runtime characteristics.